Compositional Quantitative Resource Analysis of C Programs

### 2014-06-24 paper/main.tex
minor
Q:Call is wrong

  This paper describes a novel quantitative resource analysis
  framework that tackles these challenges for C programs.  First, the
  runtime resource cost of programs is formalized with a parametric
  cost semantics for CompCert Clight.  Second, a quantitative Hoare
  logic that can be used for interactive bound development is
  implement and proved sound with respect to the cost semantics in the
  Coq Proof Assistant.  Third, an automatic amortized resource
  analysis for C programs that computes derivations in the
  quantitative Hoare logic is developed and implemented.  Both, the
  quantitative logic and the automatic amortized analysis are
  naturally compositional and can be combined to semi-automatically
  derive global resource bounds.

@@

Our new quantitative Hoare logic for CompCert Clight is a
generalization of a program logic that we developed to formally very
stack-space bounds for C programs in Coq~\cite{veristack14}.  In a
nutshell, we replace the boolean predicates (pre and post conditions)
of Hoare logic with numeric predicates that map a program state to a
non-negative number or infinity.  The intuitive meaning of a
quantitative Hoare triple $\htriple{P}{S}{Q}$ that $C$ consumes $P(\state)$
resources when executed in starting state $\state$, and that $Q(\state')$
resources are left if $C$ terminates in state $\state'$.  Infinity
corresponds to \emph{false} in classic Hoare logic.  Our quantitative
logic is proved sound with respect to a cost semantic for Clight that
accounts a constant cost for each evaluation step.  This constant cost
is parametric in a resource metric that can be defined by the user.
Our shallow embedding in Coq enables the derivation of both, bounds
that are parametric in the cost metric or bounds for one specific
metric.

@@

A program state $\state = (\theta, \gamma)$ composed of two maps from
variable names to integers. The first map $\theta : \text{Locals} \to
\Z$ assigns integers to local variables of a function and the second
map $\gamma : \text{Globals} \to \Z$ gives values to global variables
of the program.
%
The evaluation function \evalE{\cdot}{} that maps an expression $e \in
E$ to a value $\evalE{e}{\state} \in \Z$ in program state $\state$.

@@

% The continuation-based semantics of a Clight requires that we
% distinguish pre- and postconditions in the logic to account for
% different possible ways to exit a block of code.  This is approach is
% standard in Hoare logics and followed for instance in Appel's
% separation logic for Clight~\cite{AppelLogic}.  Our postconditions
% $$
% Q = (Q^s, Q^b, Q^r) : \Assn \times \Assn \times (\Val \to \Assn)
% $$
% provide one assertion $Q^s$ for the case in which the block is exited by
% fall through, one assertion $Q^b$ if the block is exited by a \code{break}, and a
% function $Q^r$ from values to assertions in case the block is exited by a
% \code{return}.  The function argument in the last case represents the return
% value and the intended meaning is that the resulting assertion is
% guaranteed to hold for every return value.

@@

\begin{figure*}[t]
\begin{mathpar}
%
\Rule{Q:Skip}
{}
{ \Gamma; B; R; Q \vdash \code{skip} \dashv Q; \Gamma }
%
\and \Rule{Q:Break}
{}
{ \Gamma; B; R; B + M_b \vdash \code{break} \dashv Q'; \Gamma' }
%
\and \Rule{Q:Return}
{}
{ \Gamma; B; R; R \, x + M_r \vdash \code{return}~x \dashv Q'; \Gamma' }
%
\and \Rule{Q:Assert}
{}
{ \Gamma; B; R; Q + M_a \vdash \code{assert}~e \dashv Q; \Gamma, e }
%
\and \Rule{Q:Loop}
{ \Gamma; Q'; R; Q \vdash S \dashv M_l + Q; \Gamma }
{ \Gamma; B; R; Q \vdash \code{loop}~S \dashv Q'; \code{inv}_S~\Gamma }
%
\and \Rule{Q:Set}
{ q'_{xy}, q'_{yx} \in \mathbb Q_0^+
\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
}
{ \Gamma[x/y]; B; R; Q + M_s \vdash x \gets y \dashv Q'; \Gamma }
%
\and \Rule{Q:IncrPos}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\}
\\\\ \textstyle q'_{0y} = q_{0y}
      + \sum_{u \in \mathcal U} q_{xu}
      - \sum_{v \not\in \mathcal U} q_{vx}
}
{ \textstyle
  \Gamma[x/x + y]; B; R; Q + M_s \vdash x \gets x + y \dashv Q'; \Gamma
}
%
\and \Rule{Q:IncrNeg}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
\\\\ \textstyle q'_{y0} = q_{y0}
      + \sum_{u \in \mathcal U} q_{ux}
      - \sum_{v \not\in \mathcal U} q_{xv}
}
{ \textstyle
  \Gamma[x/x - y]; B; R; Q + M_s \vdash x \gets x - y \dashv Q'; \Gamma
}
%
\and \Rule{Q:Incr}
{ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
}
{ \textstyle
  \Gamma[x/x \pm y]; B; R; Q + M_s \vdash x \gets x \pm y \dashv Q'; \Gamma
}
%
\and \Rule{Q:If}
{ \Gamma, e; B; R; Q - M_c^1 \vdash S_1 \dashv Q'; \Gamma'
\\ \Gamma, \neg e; B; R; Q - M_c^2 \vdash S_2 \dashv Q'; \Gamma'
}
{ \Gamma; B; R; Q \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv Q'; \Gamma' }
%
\and\Rule{Q:Seq}
{  \Gamma; B; R; Q - M_s^1 \vdash S_1 \dashv Q'; \Gamma'
\\ \Gamma'; B; R; Q' - M_s^2 \vdash S_2 \dashv Q''; \Gamma''
}
{ \Gamma; B; R; Q \vdash S_1;S_2 \dashv Q''; \Gamma'' }
%
\and\Rule{Q:Call}
{ (\vec y, s, \Gamma_f, Q_f, Q_f', \Gamma_f') \in \Delta(f)
\\ P = Q_f \Uparrow_{[\vec y = \vec x]}
\\ P' = Q'_f \Uparrow_{[r = s]}
\\ \forall i \in \ell \,\setminus \{r\} .\, q_i = q_i'
\\ \forall i \not\in \ell \,\setminus \{r\} .(q_i = p_i \land q'_i = p'_i)
\\ c \in \mathbb Q_0^+
}
{ \Gamma_f[\vec y / \vec x] \land \Gamma_\ell; B; R; Q + c
  \vdash r \gets f(\vec x) \dashv
  Q' + c; \Gamma_f'[s/r] \land \Gamma_\ell
}
%
\and \Rule{Q:Weak}
{ \Gamma_1 \models \Gamma_2
\\ Q_1 \succeq_\Gamma Q_2
\\ \Gamma_2; B; R; Q_2 \vdash S \dashv Q'_2; \Gamma'_2
\\ Q'_2 \succeq_\Gamma Q'_1
\\ \Gamma'_2 \models \Gamma'_1
}
{ \Gamma_1; B; R; Q_1 \vdash S \dashv Q'_1; \Gamma'_1 }
%
\and \Rule{Relax}
{ \mathcal L = \{ xy \mid \exists l_{xy},~\Gamma \models l_{xy} \le |\inter x y| \}
\and \mathcal U = \{ xy \mid \exists u_{xy},~\Gamma \models |\inter x y| \le u_{xy} \}
\\ \forall i \in \mathcal U.\, q'_i \ge q_i - p_i
\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
\\ \forall i \not\in \mathcal U \cup \mathcal L.\, q'_i \ge q_i
}
{ Q' + \sum_{i \in \mathcal U} u_i p_i - \sum_{i \in \mathcal L} l_i p_i  \succeq_\Gamma Q }
%
\end{mathpar}
\caption{Abstract proof rules of the quantitative analysis}
\label{fig:proof}
\end{figure*}

>>>>

  This paper describes a novel quantitative resource analysis
  framework that tackles these challenges for C programs.  The
  analysis framework consists of three parts.  First, the runtime
  resource cost of programs is formalized with a parametric cost
  semantics for CompCert Clight.  Second, a quantitative Hoare logic
  that can be used for interactive bound development is implement and
  proved sound with respect to the cost semantics in the Coq Proof
  Assistant.  Third, an automatic amortized resource analysis for C
  programs that computes derivations in the quantitative Hoare logic
  is developed and implemented.  Both, the quantitative logic and the
  automatic amortized analysis are naturally compositional and can be
  combined to semi-automatically derive global resource bounds.

@@

Our new quantitative Hoare logic for CompCert Clight is a
generalization of a program logic that we developed to formally very
stack-space bounds for C programs in Coq~\cite{veristack14}.  In a
nutshell, we replace the boolean predicates (pre and post conditions)
of Hoare logic with numeric predicates that map a program state to a
non-negative number or infinity.  The intuitive meaning of a
quantitative Hoare triple $\htriple{P}{S}{Q}$ is that $S$ consumes
less then $P(\state)$ resources when executed in starting state
$\state$, and that $Q(\state')$ resources are left if $S$ terminates
in state $\state'$.  Infinity corresponds to \emph{false} in classic
Hoare logic.  Our quantitative logic is proved sound with respect to a
cost semantic for Clight that accounts a constant cost for each
evaluation step.  This constant cost is parametric in a resource
metric that can be defined by the user.  Our shallow embedding in Coq
enables the derivation of both, bounds that are parametric in the cost
metric or bounds for one specific metric.

@@

A program state $\state = (\theta, \gamma)$ composed of two maps from
variable names to integers. The first map $\theta : \text{Locals} \to
\Z$ assigns integers to local variables of a function and the second
map $\gamma : \text{Globals} \to \Z$ gives values to global variables
of the program.  In this article, we assume that all values are
integers but in the implementation we support all data types of
Clight.
%
The evaluation function \evalE{\cdot}{} that maps an expression $e \in
E$ to a value $\evalE{e}{\state} \in \Z$ in program state $\state$.

@@

Due to \code{break} and \code{return} statements of Clight, there are
different possible ways to exit a block of code.  We also have to keep
track of the resource specifications of functions.  To account for
this in the logic, our quantitative Hoare triples have the form
$$
\Delta; B; R \vdash
\htriple
  { Q }
  { S }
  { Q' } \; .
$$
The triple $\htriple{ Q }{ S }{ Q' }$ consists of a statement $S$ and
two assertions $Q,Q' : \Assn$.  It corresponds to triples in classic
Hoare logic and the intuitive meaning is as follows.  If $S$ is
executed with starting state $\state$, the empty continuation
$\Kstop$, and at least $P(\state)$ resources available then the
evaluation does not run out of resources and there are at least
$Q(\state')$ resources left if the evaluation terminates in state
$\state'$.  However, the assertion $B : \Assn$ provides the
postcondition for the case in which the code block $S$ is exited by a
\code{break} statement.  So if the execution is terminated in state
$\state'$ with a \code{break} then $B(\state')$ resources are
available.  Similarly, $R : \Z \to \Assn$ is the postcondition for the
case in which the code block $S$ is exited by a \code{return x}
statement.  The integer argument of $R$ is the return value.
Finally, the function context 
$$\Delta {:} \text{Funs} \to \Z^*
\\
((\Val \times \Mem) {\to} \N \cup \{\infty\}) {\times} ((\Val \times \Mem) {\to} \N {\cup} \{\infty\})$$

% The continuation-based semantics of a Clight requires that we
% distinguish pre- and postconditions in the   This is approach is
% standard in Hoare logics and followed for instance in Appel's
% separation logic for Clight~\cite{AppelLogic}.  Our postconditions
% $$
% Q = (Q^s, Q^b, Q^r) : \Assn \times \Assn \times (\Val \to \Assn)
% $$
% provide one assertion $Q^s$ for the case in which the block is exited by
% fall through, one assertion $Q^b$ if the block is exited by a \code{break}, and a
% function $Q^r$ from values to assertions in case the block is exited by a
% \code{return}.  The function argument in the last case represents the return
% value and the intended meaning is that the resulting assertion is
% guaranteed to hold for every return value.

@@

\begin{figure*}[t]
\begin{mathpar}
%
\Rule{Q:Skip}
{}
{ \Gamma; B; R; Q \vdash \code{skip} \dashv Q; \Gamma }
%
\and \Rule{Q:Break}
{}
{ \Gamma; B; R; B + M_b \vdash \code{break} \dashv Q'; \Gamma' }
%
\and \Rule{Q:Return}
{}
{ \Gamma; B; R; R \, x + M_r \vdash \code{return}~x \dashv Q'; \Gamma' }
%
\and \Rule{Q:Assert}
{}
{ \Gamma; B; R; Q + M_a \vdash \code{assert}~e \dashv Q; \Gamma, e }
%
\and \Rule{Q:Loop}
{ \Gamma; Q'; R; Q \vdash S \dashv M_l + Q; \Gamma }
{ \Gamma; B; R; Q \vdash \code{loop}~S \dashv Q'; \code{inv}_S~\Gamma }
%
\and \Rule{Q:Set}
{ q'_{xy}, q'_{yx} \in \mathbb Q_0^+
\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
}
{ \Gamma[x/y]; B; R; Q + M_s \vdash x \gets y \dashv Q'; \Gamma }
%
\and \Rule{Q:IncrPos}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\}
\\\\ \textstyle q'_{0y} = q_{0y}
      + \sum_{u \in \mathcal U} q_{xu}
      - \sum_{v \not\in \mathcal U} q_{vx}
}
{ \textstyle
  \Gamma[x/x + y]; B; R; Q + M_s \vdash x \gets x + y \dashv Q'; \Gamma
}
%
\and \Rule{Q:IncrNeg}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
\\\\ \textstyle q'_{y0} = q_{y0}
      + \sum_{u \in \mathcal U} q_{ux}
      - \sum_{v \not\in \mathcal U} q_{xv}
}
{ \textstyle
  \Gamma[x/x - y]; B; R; Q + M_s \vdash x \gets x - y \dashv Q'; \Gamma
}
%
\and \Rule{Q:Incr}
{ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
}
{ \textstyle
  \Gamma[x/x \pm y]; B; R; Q + M_s \vdash x \gets x \pm y \dashv Q'; \Gamma
}
%
\and \Rule{Q:If}
{ \Gamma, e; B; R; Q - M_c^1 \vdash S_1 \dashv Q'; \Gamma'
\\ \Gamma, \neg e; B; R; Q - M_c^2 \vdash S_2 \dashv Q'; \Gamma'
}
{ \Gamma; B; R; Q \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv Q'; \Gamma' }
%
\and\Rule{Q:Seq}
{  \Gamma; B; R; Q - M_s^1 \vdash S_1 \dashv Q'; \Gamma'
\\ \Gamma'; B; R; Q' - M_s^2 \vdash S_2 \dashv Q''; \Gamma''
}
{ \Gamma; B; R; Q \vdash S_1;S_2 \dashv Q''; \Gamma'' }
%
\and\Rule{Q:Call}
{ (\vec y, s, \Gamma_f, P, P', \Gamma_f') \in \Delta(f)
\\ \ell = \text{Locals}(Q)
\\ R = P[\vec y / \vec x]
\\ \forall i \in \dom{R} . \, q_i = r_i
\\ R' = P'[s/r]
\\ \forall i \in \dom{R'} .\, q'_i = r'_i
\\ \forall i \not\in \dom{R'} .\, q'_i = 0
\\ \forall i \in \ell . \, q_i = q_i'
% \\ P' = Q'_f \Uparrow_{[r = s]}
% \\ \forall i \in \ell \,\setminus \{r\} .\, q_i = q_i'
% \\ \forall i \not\in \ell \,\setminus \{r\} .(q_i = p_i \land q'_i = p'_i)
% \\ c \in \mathbb Q_0^+
}
{ \Gamma_f[\vec y / \vec x] \land \Gamma_\ell; B; R; Q + c
  \vdash r \gets f(\vec x) \dashv
  Q' + c; \Gamma_f'[s/r] \land \Gamma_\ell
}
%
\and \Rule{Q:Weak}
{ \Gamma_1 \models \Gamma_2
\\ Q_1 \succeq_\Gamma Q_2
\\ \Gamma_2; B; R; Q_2 \vdash S \dashv Q'_2; \Gamma'_2
\\ Q'_2 \succeq_\Gamma Q'_1
\\ \Gamma'_2 \models \Gamma'_1
}
{ \Gamma_1; B; R; Q_1 \vdash S \dashv Q'_1; \Gamma'_1 }
%
\and \Rule{Relax}
{ \mathcal L = \{ xy \mid \exists l_{xy},~\Gamma \models l_{xy} \le |\inter x y| \}
\and \mathcal U = \{ xy \mid \exists u_{xy},~\Gamma \models |\inter x y| \le u_{xy} \}
\\ \forall i \in \mathcal U.\, q'_i \ge q_i - p_i
\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
\\ \forall i \not\in \mathcal U \cup \mathcal L.\, q'_i \ge q_i
}
{ Q' + \sum_{i \in \mathcal U} u_i p_i - \sum_{i \in \mathcal L} l_i p_i  \succeq_\Gamma Q }
%
\end{mathpar}
\caption{Abstract proof rules of the quantitative analysis}
\label{fig:proof}
\end{figure*}
